The Soulgem Paper

### 2011-09-24 soulgem.tex
Second draft of Related Work (which now comes before Theoretical Framework).

@@

Sociologists and organizational scientists have explored the topic of professional and social roles for many years. Biddle \cite{Biddle1986} provides a good point of entry into the theoretical work in this area, as it teases out the confusion and the overloading of the concept of roles. His overview covers five main variations of role theory: \emph{functional} role theory (roles are functions in a stable social system), \emph{symbolic interactionist} role theory (roles evolve through social interaction, norms only provide broad imperatives), \emph{structural} role theory (roles are patterns of behaviour shared among people in similar positions in the social structure), \emph{organizational} role theory (which assumes that roles are mainly set by social positions in preplanned, rational settings), and \emph{cognitive} role theory (which focuses on determining which conditions give rise to expectations, how to measure expectations, and what impact do they have on social conduct). Our theoretical framework above draws mainly from the symbolic interactionist theory as described by Biddle, but although there are significant philosophical and methodological differences between role theories, they are compatible on many essential points. As Biddle points out, ``most versions of role theory presume that expectations are the major generators of roles, that expectations are learned through experience, and that persons are aware of the expectations they hold.''

Although there are roles that everyone in an organization is expected to enact (such as being a good team member), the roles that are useful in most analyses are those that allow us to differentiate between people. As Ashforth and Mael \cite{Ashforth1989} point out, we define ourselves partly in opposition to what others do. These distinctions generate both conflicts and opportunities for the members of a social group. Goode \cite{Goode1960} talks about the stresses related with role strain, argues that in general an individual's role obligations are overdemanding, and discusses coping mechanisms commonly used. In contrast, Sieber \cite{Sieber1974} claims that the additional demands caused by accumulating roles also bring rewards, and that these rewards may well compensate for the corresponding strain in many cases.

Researchers discuss roles frequently, in both prescriptive and descriptive efforts, although prescriptions appear to outnumber descriptions. Chief Programmer Teams \cite{Baker1972} are an early example in our literature; they are modelled after surgical teams, with a Chief Programmer assisted by a group of professionals. Process definitions tend to include role prescriptions: Scrum \cite{Schwaber2001}, for instance, calls for a Scrum Master, a Product Owner, and a number of Developers. Practices also prescribe roles, but in a more minute detail and, usually, for a shorter term. Thus in a Fagan inspection \cite{Fagan1976} we have a Moderator, a Reader, a Coder, and several Reviewers.

Recent work in this community regularly discusses roles in interesting ways, but without a solid or consistent theoretical framework. Hoda \emph{et al.}\ \cite{Hoda2010} identified six ``informal roles'' that team members adopt in self-organizing teams. These roles (Mentor, Co-ordinator, Translator, Champion, Promoter, and Terminator) are separate from the people and positions that enact them (Agile Coaches, Developers, Business Analysts), a distinction we advocate for in this paper. But several of the ``roles'' they identify (such as Champion and Promoter) are too fine-grained, and they fall under the category of ``expectations'' from our perspective; expectations that are bundled with others to conform a richer, multi-faceted role. Moe \emph{et al.} \cite{Moe2010} use Dickinson and McIntyre's \cite{Dickinson1997} teamwork model to provide such a set of expectations for members of self-managing teams. While useful, this model is overly generic for our purposes, as it does not allow us to study how or why differences in expectations and specializations within teams arise.

Other researchers have attempted to study roles structurally, as they arise by social network interactions. For instance, Toral \emph{et al.}\ \cite{Toral2010} studied ``middle-men'' or ``brokers,'' and Ricca and Marchetto \cite{Ricca2010} examine the role of ``heroes,'' in open source projects. Their structural analysis allows them to scale to large projects and electronic repositories at the expense of detail. To our knowledge, more sophisticated analyses of roles in social networks, such as the blockmodelling work of Breiger and Pattison \cite{Breiger1986}, has not been carried out in our domain.

We can also find rich descriptions of some roles in software development. Martin \emph{et al.}\ \cite{Martin2007} report a detailed picture of the testing role carried out by the developers in a software organization, and of the reasons for which they deviate from the testing prescriptions in the literature. F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a rich report of the consequences of an attempted role evolution in a case in which developers attempted to rotate the job of customer support.

>>>>

Sociologists and organizational scientists have explored the topic of professional and social roles for a long time. Biddle \cite{Biddle1986} provides a good point of entry into the theoretical work in this area, as it teases out the confusion and the overloading of the concept of roles. His overview covers five main variations of theoretical thinking: \emph{functional} role theory (roles are functions in a stable social system), \emph{symbolic interactionist} role theory (roles evolve through social interaction, norms only provide broad imperatives), \emph{structural} role theory (roles are patterns of behaviour shared among people in similar positions in the social structure), \emph{organizational} role theory (roles are mainly set by social positions in preplanned, rational settings), and \emph{cognitive} role theory (which focuses on determining which conditions give rise to expectations, how to measure expectations, and what impact do they have on social conduct). Although there are significant philosophical and methodological differences between role theories, they are compatible on many essential points. As Biddle points out, ``most versions of role theory presume that expectations are the major generators of roles, that expectations are learned through experience, and that persons are aware of the expectations they hold.''

% Could take this para out if we need space.
Although there are roles that everyone in an organization is expected to enact, the roles that are useful in most analyses are those that allow us to differentiate between people. As Ashforth and Mael \cite{Ashforth1989} point out, we define ourselves partly in opposition to what others do. These distinctions generate both conflicts and opportunities for the members of a social group.

% And if we need *more* space, this is only tangentially relevant.
Goode \cite{Goode1960} talks about the stresses related with role strain, argues that in general an individual's role obligations are over-demanding, and discusses coping mechanisms commonly used. In contrast, Sieber \cite{Sieber1974} claims that the additional demands caused by accumulating roles also bring rewards, and that these rewards may well compensate for the corresponding strain in many cases.

In the software engineering literature, researchers discuss roles frequently, in both prescriptive and descriptive efforts. Chief Programmer Teams \cite{Baker1972} are an early example of a prescriptive approach. Process definitions tend to include role prescriptions: Scrum \cite{Schwaber2001}, for instance, calls for a Scrum Master, a Product Owner, and a number of developers. Practices also prescribe roles, but in a more minute detail and, usually, for a shorter term. Thus in a Fagan inspection \cite{Fagan1976} we have a Moderator, a Reader, a Coder, and several Reviewers.

Recent work regularly discusses roles in interesting ways, but without a solid or consistent theoretical framework. Hoda \emph{et al.}\ \cite{Hoda2010} identified six ``informal roles'' that team members adopt in self-organizing teams. The authors distinguish between the roles and the people and positions that enact them, which is a distinction we advocate for in this paper. But several of the roles they identify (such as Champion and Promoter) are too fine-grained, and they fall under the category of expectations from our perspective, expectations that are bundled with others to conform a richer, multi-faceted role, as we will discuss in the next section.

Moe \emph{et al.} \cite{Moe2010} use Dickinson and McIntyre's \cite{Dickinson1997} teamwork model to provide such a set of expectations for members of self-managing teams. This is certainly useful, although it cannot be used as a basis for our work, as it does not allow us to study how or why differences in expectations and specializations within teams arise.

Other researchers have attempted to study roles structurally, as they arise by social network interactions. For instance, Toral \emph{et al.}\ \cite{Toral2010} studied ``middle-men'' or ``brokers,'' and Ricca and Marchetto \cite{Ricca2010} examine the role of ``heroes,'' in open source projects. Their structural analysis allows them to scale to large projects and electronic repositories, at the expense of detail. To our knowledge, more sophisticated analyses of roles from a structural role theory perspective, such as the sociological blockmodelling work of Breiger and Pattison \cite{Breiger1986}, has not been carried out in our domain.

We can also find rich descriptions of some roles in software development. Martin \emph{et al.}\ \cite{Martin2007} report a detailed picture of the testing role carried out by the developers in a software organization, and of the reasons for which they deviate from the testing prescriptions in the literature. F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a comprehensive report of the consequences of role evolution in a case in which developers attempted to rotate the responsibilities of customer support.

@@
